Model Search in Relational Domains

### 2016-10-22 algorithm.tex
section 5.1

\section{Abstract Conflict Driven Clause Learning}
%
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKw{return}{return}
\SetKwRepeat{Do}{do}{while}
%\SetKwFunction{assume}{assume}
%\SetKwFunction{isf}{isFeasible}
\SetKwData{conflict}{conflict}
\SetKwData{safe}{safe}
\SetKwData{sat}{sat}
\SetKwData{unsafe}{unsafe}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{A program in the form of a set of abstract transformers $\abstransset$.}
\Output{The status \safe or \unsafe, and a counterexample if \unsafe.}
$\trail \leftarrow \langle\rangle$ \;
$\reasons \leftarrow []$ \;
$\mathit{result} \leftarrow \deduce_{\propheur}(\abstransset,\trail,\reasons)$ \;
\lIf{$\mathit{result}$ = \conflict} {
  \return \safe}
\While{$true$} 
{
\lIf{$\mathit{result}$ = \sat} {
  \return (\unsafe,$\abs(\trail)$)}
  $\decisionvar \leftarrow \decide_{\decheur}(\abs(\trail))$ \;
  $\trail \leftarrow \trail \cdot \decisionvar$ \; 
  $\reasons[|\trail|] \leftarrow \top$ \;
  $\mathit{result} \leftarrow \deduce_{\propheur}(\abstransset,\trail,\reasons)$\;
  \Do{$\mathit{result} = \conflict$} {
    \lIf{$\neg \analyzeconflict_{\confheur}(\abstransset,\trail,\reasons)$} {
      \return \safe
    }
    $\mathit{result} \leftarrow \deduce_{\propheur}(\abstransset,\trail,\reasons)$ \;
  }
}
\end{small}
\caption{Abstract Conflict Driven Clause Learning $ACDCL_{\propheur,\decheur,\confheur}(\abstransset)$ \label{Alg:acdcl}}
\end{algorithm2e}
%
%\subsection{Algorithm for Abstract Conflict Driven Clause Learning}
Algorithm~\ref{Alg:acdcl} presents an overview of the ACDCL algorithm.
The input to the algorithm is a program in the form of the set of its abstract transformers $\abstransset$ w.r.t.\ an abstract domain~$\domain$. The algorithm is parametrised by propagation, decision, and conflict analysis
heuristics.  
%\rmcmt{Approximation of the concrete transformers in 
%$\abstransset$ are typically available in abstract domain in the 
%form of strongest-post condition or weakest pre-condition. } 
It maintains a propagation
trail $\trail$ and a reason trail $\reasons$.
The propagation trail is an initially empty sequence of meet
irreducibles inferred by the abstract model search phase (deductions
and decisions). The reason trail maps the elements of the
propagation trail to transformers $\abstrans\in\abstransset$ that were used to
derive them. The \emph{abstract value} $\abs(\trail)$
corresponding to the propagation trail $\trail$ is the
conjunction of the meet irreducibles on the trail:
$\abs(\trail)=\bigsqcap_{m \in \trail}m$ with
$\abs(\trail)=\top$ if $\trail$ is the empty
sequence $\langle\rangle$.

The procedure $\deduce$ computes a greatest fixed point over the
transformers in $\abstransset$ that refines the abstract value,
similar to the BCP step in SAT solvers.  If the result of $\deduce$
returns \textsf{conflict} ($\bot$), the algorithm terminates with
\textsf{safe}.  Else, the analysis enters into the while($\true$) loop
and makes a new decision by a call to $\decide$ which returns a new
meet irreducible $\decisionvar$.
%
%\pscmt{what happens if we cannot make any decision any more?
%  -- then we should either have deduced $\bot$ or $\abs(\trail)$ must
%  be $\gamma$-complete} 
%
We concatenate $\decisionvar$ to the
trail $\trail$.  The decision $\decisionvar$ therefore refines the current
abstract value $\abs(\trail)$ represented by the trail.
% WE ExPLAIN THAT LATER
%However, $\decisionvar$ must be consistent with the state of the
%solver, i.e. $\abs(\trail\cdot \decisionvar)\neq \bot$.
%
For example, a decision in the interval domain restricts the range of 
intervals for variables.
%
We set the corresponding entry in the reason trail $\reasons$ to $\top$
to mark it as a decision.
% so the analysis jumps under a 
%greatest fixed-point \pscmt{???}. Note that the widening operation in abstract interpretation 
%jumps above a least fixed-point, so decisions can be viewed as 
%{\em dual widening} \pscmt{what do we learn?}.

The procedure $\deduce$ is called next to infer new meet irreducibles
taking into account the current decision.  The model search phase
alternates between the decision and deduction until $\deduce$ returns
either \textsf{sat} or \textsf{conflict}.  In the former case, we have
found an abstract value that represents models of $\formula$, which
are counterexamples to the safety problem, and thus we return
\textsf{unsafe}.
%
% has found an abstract value that represents a set of models of $\formula$ or the
% deduction leads to a \textsf{conflict}.  Checking whether the abstract
% model concretises to a set of models of $\formula$; this
% corresponds to a $\gamma$-completeness~\cite{dhk2013-popl} check in
% abstract interpretation.  If $v$ is $\gamma$-complete, then it cannot
% be refined further.  Thus, the algorithm returns the abstract model
% $v$, which is a set of concrete models, and terminates with
% \textsf{unsafe} or the current abstraction is insufficient to
% determine the satisfiability of $\formula$.

In the latter case the algorithm enters in the $\analyzeconflict$
phase to learn the reason for the conflict.   There can be multiple
incomparable reasons for conflict.
% \pscmt{do you have to explain this here?} -- based on the choice of Unique
%Implication Point (UIP)~\cite{cdcl}.  
ACDCL heuristically chooses one reason and learns it by adding at as
an abstract transformer to $\abstransset$. Then it backtracks by
removing the tail of the $\trail$ up to a point where it becomes
consistent, and continues making deductions from there.
If $\analyzeconflict$ returns $\false$ then $\formula$ is unsatisfiable
and we return  \textsf{safe}.

>>>>

\section{Abstract Conflict Driven Clause Learning}
%
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKw{return}{return}
\SetKwRepeat{Do}{do}{while}
%\SetKwFunction{assume}{assume}
%\SetKwFunction{isf}{isFeasible}
\SetKwData{conflict}{conflict}
\SetKwData{safe}{safe}
\SetKwData{sat}{sat}
\SetKwData{unsafe}{unsafe}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{A program in the form of a set of abstract transformers $\abstransset$.}
\Output{The status \safe or \unsafe, \rmcmt{and a counterexample if \unsafe.}}
$\trail \leftarrow \langle\rangle$ \;
$\reasons \leftarrow []$ \;
$\mathit{result} \leftarrow \deduce_{\propheur}(\abstransset,\trail,\reasons)$ \;
\lIf{$\mathit{result}$ = \conflict} {
  \return \safe}
\While{$true$} 
{
\lIf{$\mathit{result}$ = \sat} {
  \return (\unsafe,$\abs(\trail)$)}
  $\decisionvar \leftarrow \decide_{\decheur}(\abs(\trail))$ \;
  $\trail \leftarrow \trail \cdot \decisionvar$ \; 
  $\reasons[|\trail|] \leftarrow \top$ \;
  $\mathit{result} \leftarrow \deduce_{\propheur}(\abstransset,\trail,\reasons)$\;
  \Do{$\mathit{result} = \conflict$} {
    \lIf{$\neg \analyzeconflict_{\confheur}(\abstransset,\trail,\reasons)$} {
      \return \safe
    }
    $\mathit{result} \leftarrow \deduce_{\propheur}(\abstransset,\trail,\reasons)$ \;
  }
}
\end{small}
\caption{Abstract Conflict Driven Clause Learning $ACDCL_{\propheur,\decheur,\confheur}(\abstransset)$ \label{Alg:acdcl}}
\end{algorithm2e}
%
%\subsection{Algorithm for Abstract Conflict Driven Clause Learning}
Algorithm~\ref{Alg:acdcl} presents an overview of the ACDCL algorithm.
The input to the algorithm is a program in the form of the set of its 
abstract transformers $\abstransset$ w.r.t.\ an abstract domain~$\domain$.  
The algorithm is parametrised by propagation \rmcmt{$(\propheur)$}, decision
$(\decheur)$, and conflict analysis heuristics $(\confheur)$.  
%\rmcmt{Approximation of the concrete transformers in 
%$\abstransset$ are typically available in abstract domain in the 
%form of strongest-post condition or weakest pre-condition. } 
The algorithm begins with an empty propagation trail $\trail$ and 
a reason trail $\reasons$.
%The propagation trail is initialized with an empty sequence.  
The propagation trail stores all meet irreducibles inferred by 
the abstract model search phase (deductions and decisions).  
The reason trail maps the elements of the propagation trail to 
transformers $\abstrans\in\abstransset$ that were used to
derive them. 
%
\begin{definition} 
The \emph{abstract value} $\abs(\trail)$ corresponding to 
the propagation trail $\trail$ is the conjunction of the 
meet irreducibles on the trail:
$\abs(\trail)=\bigsqcap_{m \in \trail}m$ with
$\abs(\trail)=\top$ if $\trail$ is the empty sequence $\langle\rangle$.
\end{definition}
%
The procedure $\deduce$ computes a greatest fixed point over the
transformers in $\abstransset$ that refines the abstract value,
similar to the BCP step in SAT solvers.  If the result of $\deduce$
returns \textsf{conflict} ($\bot$), the algorithm terminates with
\textsf{safe}.  Else, the analysis enters into the while($\true$) loop
and makes a new decision by a call to $\decide$ which returns a new
meet irreducible $\decisionvar$.
%
%\pscmt{what happens if we cannot make any decision any more?
%  -- then we should either have deduced $\bot$ or $\abs(\trail)$ must
%  be $\gamma$-complete} 
%
We concatenate $\decisionvar$ to the
trail $\trail$.  The decision $\decisionvar$ therefore refines \rmcmt{not
defined} the current abstract value $\abs(\trail)$ represented by the trail. 
\rmcmt{a decision is added to the trail only if it refines the trail.}
% WE ExPLAIN THAT LATER
%However, $\decisionvar$ must be consistent with the state of the
%solver, i.e. $\abs(\trail\cdot \decisionvar)\neq \bot$.
%
For example, a decision in the interval domain restricts the range of 
intervals for variables.
%
We set the corresponding entry in the reason trail $\reasons$ to $\top$
\rmcmt{to mark it as a decision.}
% so the analysis jumps under a 
%greatest fixed-point \pscmt{???}. Note that the widening operation in abstract interpretation 
%jumps above a least fixed-point, so decisions can be viewed as 
%{\em dual widening} \pscmt{what do we learn?}.

The procedure $\deduce$ is called next to infer new meet irreducibles
taking into account the current decision.  The model search phase
alternates between the decision and deduction until $\deduce$ returns
either \textsf{sat} or \textsf{conflict}.  In the former case, we have
found an abstract value that represents models of $\formula$, which
are counterexamples to the safety problem, and thus \rmcmt{we or ACDCL} return
\textsf{unsafe}.
%
% has found an abstract value that represents a set of models of $\formula$ or the
% deduction leads to a \textsf{conflict}.  Checking whether the abstract
% model concretises to a set of models of $\formula$; this
% corresponds to a $\gamma$-completeness~\cite{dhk2013-popl} check in
% abstract interpretation.  If $v$ is $\gamma$-complete, then it cannot
% be refined further.  Thus, the algorithm returns the abstract model
% $v$, which is a set of concrete models, and terminates with
% \textsf{unsafe} or the current abstraction is insufficient to
% determine the satisfiability of $\formula$.

In the latter case the algorithm enters in the $\analyzeconflict$
phase to learn the reason for the conflict.   There can be multiple
incomparable reasons for conflict.
% \pscmt{do you have to explain this here?} -- based on the choice of Unique
%Implication Point (UIP)~\cite{cdcl}.  
ACDCL heuristically chooses one reason $\conflictset$ and learns it 
by adding it as an abstract transformer to $\abstransset$. The algorithm 
backtracks by removing the content of $\trail$ up to a point where it does not 
conflict with $\conflictset$.  ACDCL then performs deductions with the learnt 
transformer.  If $\analyzeconflict$ returns $\false$ then no further
backtracking is possible.  Thus, the $\formula$ is unsatisfiable
and \rmcmt{we or ACDCL} return \textsf{safe}.

### 2016-10-22 deductions.tex
section 5.1

%-------------------------------------------------------------------------------
\subsection{Abstract Deduction Transformers} \label{sec:abst}
%-------------------------------------------------------------------------------

To make our algorithm efficient, we have to focus the abstract
transformers on performing only the minimally necessary
work. 
%
We thus define a specialised variant of the abstract transformer to compute
deductions w.r.t.\ a given subdomain $\subdomain\subseteq \domain$,
which we call \emph{abstract deduction transformer}.
%
A subdomain contains a chosen subset of the elements in $\domain$ including $\bot$ and $\top$
\begin{equation}\label{eq:at2}
\abstrans{\domain}{\constraint}^\subdomain(\absval)=\absval\meet_\domain \alpha_\subdomain(\{\val\mid \val\in\gamma_\domain(\absval), \val\models \constraint\})
\end{equation}
For $\subdomain=\domain$, the abstract deduction transformer behaves 
in the same way as the abstract transformer defined in Eq.~\ref{eq:abstrans} 
in Section~\ref{sec:domains}.

@@

%The first advantage of this definition is that 
\rmcmt{This parameterisation of the abstract deduction transformer with a $\subdomain$ 
allow us to guide the propagation in {\em forward}, {\em backward} or {\em
multi-way} direction.}
%using our abstract transformers.
%we can control how we perform 
%
For example, for $\absval=(0\leq y \leq 1 \wedge 5\leq z)$, we have
$\abstrans{\intervals[\{x,y,z\}]}{x=y+z}^{\intervals[\{x\}]}(\absval)=\absval\meet(x\geq
6)$. For $x=y+z$, this performs a right-hand side to left-hand side propagation and
hence emulates a forward analysis, whereas
$\abstrans{\intervals[\{x,y,z\}]}{x=y+z}^{\intervals[\{y,z\}]}=\absval\meet\top$
emulates a backward analysis.

We call the propagation with arbitrary subdomains,
e.g. $\subdomain=\domain$, \emph{multi-way propagation}, which is able
to simultaneously perform forward and backward propagation.
%
Note that the restriction to a subdomain makes the transformer less
precise. Therefore, we have the property
$\abstrans{\domain}{\constraint}^\domain(\absval)\sqsubseteq
\abstrans{\domain}{\constraint}^\subdomain(\absval)$.
However, the choice of a subdomain makes the deductions more efficient.

>>>>

%-------------------------------------------------------------------------------
\subsection{Abstract Deduction Transformers} \label{sec:abst}
%-------------------------------------------------------------------------------
\Omit{
To make our algorithm efficient, we have to focus the abstract
transformers on performing only the minimally necessary
work. 
}
\rmcmt{The key considerations for building an abstract transformer are 
precision and efficiency.  An efficient transformer is less precise, 
and vice-versa.  Our propagation strategy provides a domain-dependant 
heuristic choice to control these variants.}
%
We thus define a specialised variant of the abstract transformer to compute
deductions w.r.t.\ a given subdomain $\subdomain\subseteq \domain$,
which we call \emph{abstract deduction transformer}.
%
A subdomain contains a chosen subset of the elements in $\domain$ including
$\bot$ and $\top$, \rmcmt{which forms a lattice structure}.  An abstract 
deduction transformer is formally defined as follows. 
\begin{equation}\label{eq:at2}
\abstrans{\domain}{\constraint}^\subdomain(\absval)=\absval\meet_\domain \alpha_\subdomain(\{\val\mid \val\in\gamma_\domain(\absval), \val\models \constraint\})
\end{equation}
For $\subdomain=\domain$, the abstract deduction transformer is identical 
to the abstract transformer defined in Eq.~\ref{eq:abstrans} 
in Section~\ref{sec:domains}.

@@

%The first advantage of this definition is that 
\rmcmt{
This parameterisation of the abstract deduction transformer with a subdomain 
$\subdomain$ allow us to guide the propagation in {\em forward}, {\em backward} 
or {\em multi-way} direction.
}
%using our abstract transformers.
%we can control how we perform 
%
For example, for $\absval=(0\leq y \leq 1 \wedge 5\leq z)$, we have
$\abstrans{\intervals[\{x,y,z\}]}{x=y+z}^{\intervals[\{x\}]}(\absval)=\absval\meet(x\geq
6)$. For $x=y+z$, this performs a right-hand side to left-hand side propagation and
hence emulates a forward analysis, whereas
$\abstrans{\intervals[\{x,y,z\}]}{x=y+z}^{\intervals[\{y,z\}]}=\absval\meet\top$
emulates a backward analysis.

We call the propagation with arbitrary subdomains,
e.g. $\subdomain=\domain$, \emph{multi-way propagation}, which is able
to simultaneously perform forward and backward propagation.
%
\Omit{
Note that a restriction to a subdomain makes a transformer less
precise.
}
Note that a restricted subdomain makes a transformer less precise, but more
efficient.  On the contrary, an unrestricted subdomain make a transformer more 
precise, but less efficient. Therefore, we have the property
$\abstrans{\domain}{\constraint}^\domain(\absval)\sqsubseteq
\abstrans{\domain}{\constraint}^\subdomain(\absval)$.
%However, the choice of a subdomain makes the deductions more efficient.
\rmcmt{Thus, a choice of subdomain helps to fine-tune the deduction inferred from a 
propagation phase.}
